---
title: "rtemis Intro Vignette"
author: "Efstathios (Stathis) D. Gennatas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
monofont: Hack
sansfont: Open Sans
mainfont: Open Sans
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, prompt = FALSE, comment = NA,
                      fig.width = 5, fig.height = 5)
options(list(rt.theme = "darkgrid", rt.fit.theme = "darkgrid"))
```

```{r, comment = "", results = "asis", echo = FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)
```

# Welcome to rtemis
__rtemis__ is a platform for advanced machine learning and visualization.  
To view the full online documentation and vignettes, visit the [rtemis website](https://rtemis.netlify.com)

The __rtemis__ project has three aims:   

* __Applied Data Science__: Make advanced data analysis efficient and accessible to all   
* __Machine Learning Research__: Provide a platform to develop and test novel ML algorithms   
* __Education__: Make ML concepts easy to learn by delivering bite-sized vignettes along with code and visualization


# Overview
__rtemis__ functions are divided into the following groups:  

* __Visualization__
     - Static: **_mplot3_** family (base graphics)
     - Dynamic: **_dplot3_** family ([plotly](https://plot.ly/r/))
* __Unsupervised Learning__
     - Clustering: **_u.\*_**
     - Decomposition: **_d.\*_**
* __Supervised Learning__
     - Classification, Regression, Survival Analysis: **_s.\*_**
* __Cross-Decomposition__
     - Sparse Canonical Correlation / Sparse Decomposition: **_x.\*_**
* __Meta-Models__  
     [Have been temporarily removed for updating]
     - Model Stacking: **_metaMod()_**
     - Modality Stacking: **_metaFeat()_**
     - Group-weighted Stacking: **_metaGroup()_**

# Setup

## Install the latest __rtemis__ version from GitHub
```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("egenn/rtemis")
```

This will install __rtemis__ with a minimal set of dependencies. A dependency check is run each time a function is called and will tell you if a package is missing. Install the following packages to begin with a reasonable lightweight setup:

```{r, eval = FALSE}
packages <- c("caret", "e1071", "gbm", "glmnet", "pbapply", "ranger", "rpart")
install.packages(packages)
```

## RStudio
You can run rtemis in the command line or using the IDE of your choice. RStudio is the preferred environment and can be downloaded [here](https://www.rstudio.com/)

## macOS
### Prerequisites
If you are installing on macOS, make sure you have installed:

* [Xcode](https://itunes.apple.com/us/app/xcode/id497799835?mt=12)
* [XQuartz](https://www.xquartz.org/)

__Note on R + Java on macOS:__
In order to run some __R__ packages that use __rJava__, like [bartMachine](https://cran.r-project.org/web/packages/bartMachine/index.html), you may need to add a link to `libjvm.dylib` inside your
__R__ `lib` folder as explained [here](https://support.rstudio.com/hc/en-us/community/posts/203663956/comments/249073727?community_id=200011958&fbclid=IwAR3DdrCdfHVeUyKsSpZeVxHM9bDZKo-7B9gmsgF_rN9a50ZlgzIEsifCrio#community_comment_249073727)

### Using Apple's BLAS
You can speed up matrix operations by using Apple's [Basic Linear Algebra Subprograms (BLAS)](https://developer.apple.com/documentation/accelerate/blas) instead of the default R BLAS.
At the MacOS terminal:
```{bash eval = FALSE}
cd /Library/Frameworks/R.framework/Resources/lib
ln -sf /System/Library/Frameworks/Accelerate.framework/Frameworks/vecLib.framework/Versions/Current/libBLAS.dylib libRblas.dylib
```

Restart R and check the version of BLAS in use:
```{r}
sessionInfo()
```

[Benchmarks](https://mpopov.com/blog/2019/6/4/faster-matrix-math-in-r-on-macos) suggest substantial speed gains for some operations.

## External frameworks
The following are all optional - install as needed.

### MXNet
To use MXNet (`s.MXN`), you need to install the MXNet system libraries first and then the R package. Follow instructions on the [MXNet website](https://mxnet.incubator.apache.org).  
This will first require installation of more system dependencies, which can be installed using [Brew](https://brew.sh)

### H2O
To use H2O (`d.H2OGLRM`, `s.H2ODL.R`, `s.H2OGBM.R`, `s.H2ORF`, `u.H2OKMEANS`), you will need to install H2O first. Follow instructions on the [H2O website](https://www.h2o.ai/products/h2o/).  

### Spark
To use Spark's ML framework (`s.MLGBM`, `s.MLMP`, `s.MLRF`), installation can be performed within R:
```{r sparklyr, eval = FALSE}
install.packages("sparklyr")
sparklyr::spark_install()
```

### Keras + TensorFlow
You can easily install Keras for R and the TensorFlow library:
```{r, eval = FALSE}
devtools::install_github("rstudio/keras")
library(keras)
install_keras()
```
Learn more on the [RStudio website](https://keras.rstudio.com/)


## Load the library
```{r}
library(rtemis)
```


# Visualization
Visualization is a vital part of all data analysis. __rtemis__ has support for both static and dynamic / interactive graphics.  

## Static graphics: `mplot3`
Static plotting in __rtemis__ is handled by the `mplot3` family. `mplot3` uses base R graphics to render high quality output fast.  
Selection of available functions:

```{r echo = FALSE, results = "asis"}
viz <- data.frame(Command = c("mplot3.x", "mplot3.xy",
                              "mplot3.xym", "mplot3.fit",
                              "mplot3.img", "mplot3.heatmap",
                              "mplot3.bar", "mplot3.box"),
                  Input = c("Vector(s) `x`", "Vectors `x` and `y`",
                            "Vectors `x` and `y`", "Vectors `x` and `y`",
                            "Matrix `x`", "Matrix `x`",
                            "Vector or matrix `x`", "Matrix `x`"),
                  Output = c("Index, histogram, density, and QQ-line plots",
                             "Scatter and/or line plots with optional fit line",
                             "`mplot3.xy` wrapper for true vs. estimated values",
                             "`mplot3.xy` scatter with marginal histogram and/or density plots",
                             "2D false color plots",
                             "Heatmaps with optional row and column dendrogram and colorbar",
                             "Barplots",
                             "Boxplots"))
# datatable(viz, style = 'bootstrap')
knitr::kable(viz)
# viz.xt <- xtable::xtable(viz)
# print(viz.xt, type = "html")
# print(viz)
# pander::pandoc.table(viz)
```


### `mplot3.x`
Let's create some synthetic data.  
```{r}
set.seed(4242)
x <- rnorm(100)
y <- .6 * x + 12 + rnorm(100)
```

Let's see what happens if we try to plot the `x` variable by itself:
```{r}
mplot3.x(x)
```

You get a plot of the values of `x` against their index.  
You may want to plot its histogram:
```{r}
mplot3.x(x, type = "histogram")
```

Or its density:
```{r}
mplot3.x(x, type = "density")
```

### `mplot3.xy`
Now let's plot `y` against `x`:
```{r}
mplot3.xy(x, y)
```

Let's add a fit line:
```{r}
mplot3.xy(x, y, fit = "lm")
```

Notice that the parameter `fit` tells `mplot3` to automatically estimate a fit and overplot it.  
Any __rtemis__ model can be specified for the fit. It is recommended to use `lm` for linear fits and `gam` for nonlinear fits.
Now, let's also add a confidence band defined by +/- 2 * the standard error of the fit:  
```{r}
mplot3.xy(x, y, fit = "lm", se.fit = T)
```

### `mplot3.xym`
This function adds marginal density / histogram plots to a scatter plot:  
```{r}
x <- rnorm(200)
y <- x ^ 2 - 8 + rnorm(200)
mplot3.xym(x, y)
```

### `mplot3.heatmap`
```{r}
x <- mapply(rnorm, rep(10,10))
xcor <- cor(x)
rownames(xcor) <- LETTERS[1:10]
colnames(xcor) <- 1:10
mplot3.heatmap(xcor, Rowv = TRUE, Colv = TRUE)
```

## Dynamic graphics
### Scatter: `dplot3`
`dplot3` uses the powerful [`plotly`](https://plot.ly/) API for `R` to produce interactive plots that can be viewed as HTML in a browser or exported to bitmap graphics.  
(Try these in RStudio)
```{r, eval = FALSE}
set.seed(4242)
x <- rnorm(100)
y <- .8 * x ^ 3 + 34 + rnorm(100)
dplot3(x, y, fit = "gam")
```

### Heatmap: `dplot3.heatmap`
Hover over the cells to reveal cell info. You can also drag and drop to zoom in to part of the graph (useful for larger matrices):
```{r, eval = FALSE}
x <- mapply(rnorm, rep(10,10))
x.cor <- cor(x)
dplot3.heatmap(x.cor)
```

# Unsupervised Learning: Decomposition
__rtemis__ includes a number of algorithms for decomposition / dimensionality reduction
You can get a list of all available algorithms:  
```{r}
decomSelect()
```

As a simple example, let's look the famous `iris` dataset. Note that this is to demonstrate function usage and
in no way a good example to demonstrate the effectiveness of different decomposition algorithms since the iris
dataset only consists of 4 variables.  
Let's select all variables from the iris dataset, excluding the group names:
```{r}
x <- iris[, 1:4]
```

## Non-negative Matrix Factorization (NMF)
```{r}
iris.NMF <- d.NMF(x, k = 2)
mplot3.xy(iris.NMF$projections.train[, 1], iris.NMF$projections.train[, 2], group = iris$Species,
          xlab = "1st NMF component", ylab = "2nd NMF component", main = "NMF on iris")
```

# Unsupervised Learning: Clustering
## K-means
We'll use the NMF projection estimated above and use K-means to identify 3 clusters in the data.
```{r, warning = FALSE}
iris.kmeans <- u.KMEANS(x, k = 3)
mplot3.xy(iris.NMF$projections.train[, 1], iris.NMF$projections.train[, 2],
          group = iris.kmeans$clusters.train,
          xlab = "1st NMF component", ylab = "2nd NMF component", main = "K-MEANS on iris")
```

## Regression

### Check Data
```{r synth reg data}
x <- rnormmat(500, 50, seed = 2019)
w <- rnorm(50)
y <- x %*% w + rnorm(500)
dat <- data.frame(x, y)
res <- resample(dat)
dat.train <- dat[res$Subsample_1, ]
dat.test <- dat[-res$Subsample_1, ]
```

```{r}
checkData(x)
```


### Single Model
```{r}
mod <- s.GLM(dat.train, dat.test)
```

### Crossvalidated Model
```{r}
mod <- elevate(dat, mod = "glm")
```

Use the `describe` function to get a summary in (plain) English:
```{r}
mod$describe()
```

```{r, fig.width = 5, fig.height = 5.5}
mod$plot()
```


## Classification

### Check Data
```{r}
data(Sonar, package = 'mlbench')
checkData(Sonar)
res <- resample(Sonar)
sonar.train <- Sonar[res$Subsample_1, ]
sonar.test <- Sonar[-res$Subsample_1, ]
```

### Single model
```{r}
mod <- s.RANGER(sonar.train, sonar.test)
```


### Crossvalidated Model
```{r}
mod <- elevate(Sonar)
```

```{r}
mod$describe()
```

```{r, fig.width = 5, fig.height = 5.5}
mod$plot()
```

```{r}
mod$plotROC()
```
